Speed and strength of epidemic intervention

### 2020-01-05 doc/speedstrength.tex
Autocommit (doc)
tweaks

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

@@

\section{To do}

@@

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level. 
The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

Much research has prioritized estimates of the reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, miller2012note}.

The study of measles dynamics provides a key example for which the characterization of an epidemic via the basic reproduction number has been successfully applied in controlling the disease.
One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
Since then, measles control policies have revolved around targetting this critical vaccination threshold \citep{venczel2003measles, wolfson2007has, takahashi2015reduced}.
\swp{More citations later}
\swp{I think I need to end the paragraph more braodly so that it feels like strength vs speed comparison...}

Like epidemic strength, epidemic speed measures how disease spreads in a population.
Despite the prevailing focus on the basic reproduction number in literature \swp{CITE},
using epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
Early in an outbreak, the exponential growth rate can be directly measured from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, even though the basic reproduction number can be difficult to estimate \citep{weitz2015modeling}.
In these scenarios, we expect control efforts that seek to reduce the epidemic speed to be more effective.
The West African Ebola outbreak in 2014 provides an example for this perspective: 
many researchers tried to estimate the basic reproduction number but the major strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
Strategies that relied on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
\swp{More citations later}

\swp{Eaton and Hallett?}

Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.

@@

We model disease incidence using the renewal-equation framework, a simple, flexible framework that can cover a wide range of model structures \citep{Champredon2018equivalence}.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible.
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$. 
Importantly, this means we are neglecting changes in susceptible proportion through time;
under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.

@@

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{strength}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of an average individual as a function of time since infection) \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles (\figref{constant}A). We then have:
\begin{equation}
	\hat K(\tau) = (\RR/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread \RR.

@@

We define the strength of the intervention $L$ to be $\theta = \RR/\Rhat$. It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
\end{equation}
In this more general case, we have again that the disease cannot spread when $\theta \geq \RR$.

\jd{I'm wondering if we should be diagrammatic about these decompositions. I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}
\swp{I tried}

@@

\section{Example: HIV}

@@

We model the infection kernel of the HIV as a sum of two gamma distributions:
\begin{equation}
K(\tau) = \mathcal R \left(\tsub{p}{early} \tsub{f}{early}(\tau) + (1-\tsub{p}{early}) \tsub{f}{late}(\tau) \right).
\end{equation}
The first component, $\tsub{f}{early}(\tau)$, models early HIV transmission during the acute infection stage.
We assume that $\tsub{f}{early}(\tau)$ has a mean of 3 months \citep{hollingsworth2008hiv} and a shape parameter of 3.
The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years \swp{need to cite something and/or change this value} \jd{Happy to use any value you can cite, or to help search} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/HIV.pdf}
\caption{
\textbf{The infection kernel of the HIV.}
A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
B. Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}. The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
C. Increasing the proportion of early transmission reduces the estimate of the reproduction number.
}
\figlab{example}
\end{figure}

As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
Based on the growth rate estimate ($0.452 \textrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
Further increasing the amount of early transmission reduces the mean generation time, which in turn decreases the reproduction number (\figref{example}C);
therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is more early transmission.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/condom.pdf}
\caption{
\textbf{Understanding condom intervention using strength-based decomposition.}
A. Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $L$ due to condom use is constant across the course of infection. 
B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
}
\figlab{condom}
\end{figure}

We consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level. \jd{I'm not sure we need to be worried about weller2002condom, nor about population vs.~individual level effects. Many estimates of efficacy are higher than 80\%, and it's certainly possible to imagine an intervention that involves education about condom use.}
Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
The estimated strength of such an intervention is simply the average of $L$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/test_and_treat.pdf}
\caption{
\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
A. Test-and-treat intervention $\tsub{L}{test}$ starts at 1 and saturates after about 5 years.
B. Increasing the amount of early transmission decreases the strength of an epidemic as well as the strength of test-and-treat intervention.
C. Hazard rate $\tsub{h}{test}$ of test-and-treat intervention $\tsub{L}{test}$ peaks and plateaus very quickly and reaches zero after about 5 years.
D. The amount of early transmission has no effect on the speed of an epidemic and small effect on the speed of intervention.
We assume $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau h(\sigma) d\sigma \right)$ and $h(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $12$ months and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 1/6$.
}
\figlab{test}
\end{figure}

On the other hand, consider a test-and-treat strategy in which infected individuals are identified and put to clinical care (e.g., antiretroviral therapy).
We assume that the rate of identification and removal is initially constant. Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
\jd{Can you remind me the functional form that we used?}
However, we assume removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
\jd{I'm not sure the advantage of the tsub; I kind of feel like we can communicate in context without subscripts.} (\figref{test}A).
Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength; however, it is unclear why this should be the case. \jd{Ref Eaton}

@@

This framework provides an intuitive underpinning for the originally surprising result of Eaton: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.

>>>>

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}

@@

\section*{To do}

@@

An epidemic can be described by its \emph{speed} and \emph{strength}.
The speed of an epidemic, often characterized by the exponential growth rate $r$, measures how \emph{fast} an epidemic grows, at the population level \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}. 
The strength of an epidemic, often characterized by the basic reproductive number $\RR_0$, measures how many new cases are caused by a typical \emph{individual} case in a fully susceptible population \citep{anderson1991infectious, diekmann1990definition}.
Knowing the speed and strength of an epidemic allows predictions about the course of the epidemic and the effectiveness of intervention strategies.

Much research has prioritized estimates of the basic reproductive number, because it has a threshold value, $\RR_0=1$ that determines whether a disease can invade, the level of equilibrium, and the effectiveness of control efforts.
The insight that a case must on average cause at least one new case under good conditions for a disease persist goes back $>100$ years \citep{ross1911prevention};
the idea of averaging by defining a `typical' case was formalized 30 years ago \citep{diekmann1990definition}.
$\RR_0$ is also of interest because it provides a \emph{prima facie} prediction about the total \emph{size} of an epidemic \citep{anderson1991infectious, ma2006generality, arino2007final, andreasen2011final, miller2012note}.

Here, we generalize the idea of a threshold for successful intervention by measuring an intervention's ``strength'' on the same scale as the reproductive number. 
We then show that we can likewise measure an intervention's ``speed'', and that there is a duality between the threshold $\RR_0=1$ and a corresponding minimal intervention strength, and the threshold $r=0$ and a corresponding minimal intervention speed. 
We argue that the historical primacy of $\RR_0$ over $r$ is partly artificial, and discuss cases where strength provides the better framing for practical disease questions and cases where speed does.

@@

We model disease incidence using the renewal equation, a simple, flexible framework that can cover a wide range of model structures \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
In our model, disease incidence at time $t$ is given by:
\begin{equation}
i(t) = \int K(\tau, t) i(t-\tau) d\tau.
\end{equation}
Here, $K(\tau, t)$ is the infection kernel describing how infectious we expect an individual infected $\tau$ time units ago to be in the population.
In general, $K(\tau, t)$ will depend on population characteristics that may change through time $t$ -- notably, the proportion of the population susceptible, $S(t)$.
Since we are interested in invasion and control, we will generally neglect changes in $K(\tau, t)$ through time, thus we will assume $K(\tau, t) \equiv K(\tau)$.
Importantly, this means we are neglecting changes in susceptible proportion through time (i.e., $S(t) \approx 1$);
under this assumption, the renewal equation is equivalent to the Von Foerster equations, outlined in \cite{fraser2004factors}.

@@

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR_0 g(\tau),
	\eqlab{strength}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution.
The generation intervel is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note};
therefore, the intrinsic generation-interval distribution $g(\tau)$ gives the relative infectiousness of an average individual as a function of time since infection \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the basic reproductive number $\RR_0$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles through vaccination (\figref{constant}A). We then have:
\begin{equation}
	\hat K(\tau) = (\RR_0/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR_0$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread $\RR_0$.

@@

We define the strength of the intervention $L$ to be $\theta = \RR_0/\Rhat$. It is then straightforward to show that $\theta$ is the harmonic mean of $L(\tau)$ weighted by generation-interval distribution:
\begin{equation}
	\theta = 1/\langle 1/L(\tau) \rangle_{g(\tau)}.
\end{equation}
In this more general case, we have again that the disease cannot spread when $\theta \geq \RR_0$.

@@

\section{Example: Human immunodeficiency virus (HIV)}

@@

We model the infection kernel of the HIV as a sum of two gamma distributions:
\begin{equation}
K(\tau) = \mathcal R \left(\tsub{p}{early} \tsub{f}{early}(\tau) + (1-\tsub{p}{early}) \tsub{f}{late}(\tau) \right).
\end{equation}
The first component, $\tsub{f}{early}(\tau)$, models early HIV transmission during the acute infection stage.
We assume that $\tsub{f}{early}(\tau)$ has a mean of 3 months \citep{hollingsworth2008hiv} and a shape parameter of 3.
The second component, $\tsub{f}{late}$, models HIV transmission during the asymptomatic stage and the disease stage (after progression to Acquired Immune Deficiency Syndrome (AIDS)).
We assume that $\tsub{f}{late}(\tau)$ has a mean of 10 years to reflect its long incubation period \citep{brookmeyer1989censoring, nishiura2019estimating} and a shape parameter of 2 (this vaguely approximates the wide generation-interval distribution of the HIV \citep{fraser2004factors}).
Finally, $\tsub{p}{early}$ is the proportion of early HIV transmission.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/HIV.pdf}
\caption{
\textbf{The infection kernel of the HIV.}
A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
B. Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}. The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
C. Increase in the estimate of the amount of early transmission reduces the estimate of the reproduction number.
}
\figlab{example}
\end{figure}

As our baseline scenario (\figref{example}A), we assume that early transmission is responsible for 23\% of the total transmission \citep{hayes2006amplified}.
Based on the growth rate estimate ($0.452\,\mathrm{year}^{-1}$; \figref{example}B) from HIV prevalence in pregnant women in South Africa \citep{barron2013eliminating}, our baseline reproduction number is $3.17$.
Increase in the estimate of the amount of early transmission shortens the mean generation time, which in turn decreases the reproduction number (\figref{example}C).
Therefore, we expect stronger (but not necessarily faster) intervention to be required in order to control the disease when there is less early transmission.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/condom.pdf}
\caption{
\textbf{Understanding condom intervention using strength-based decomposition.}
A. Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $\tsub{L}{condom}$ due to condom use is constant across the course of infection.
B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
}
\figlab{condom}
\end{figure}

First, we consider a condom intervention that reduces HIV transmission by approximately 75\% at the population level.
Assuming that condoms act as a physical barrier, and that condom use will on average remain roughly constant through time, it is reasonable to model the proportional reduction in transmission due to condom use as constant across the course of infection: $\tsub{L}{condom} = 1/(1-0.75) = 4$  (\figref{condom}A).
The estimated strength of such an intervention is simply the average of $\tsub{L}{condom}$: $\theta=4$, whereas the estimated strength of the epidemic increases as the proportion of early transmission decreases (\figref{condom}B).
Thus, the predicted effectiveness of the condom intervention will depend strongly on our estimate of the importance of early transmission: if early transmission is low, we expect disease spread to be too strong to be controlled completely by our intervention.

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/test_and_treat.pdf}
\caption{
\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
A. The effectiveness of test-and-treat intervention is thought to increase exponentially until it saturates after about 5 years.
B. Increase in the estimated amount of early transmission decreases the estimated strength of an epidemic as well as the estimated strength of test-and-treat intervention.
C. The hazard rate of test-and-treat intervention stays roughly constant for the first few years and gradually decreases until it reaches zero after about 5 years.
D. The estimated amount of early transmission does not affect the observed speed of an epidemic but affects the estimated speed of intervention.
Test-and-treat intervention is modeled phenomenologically: $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau \tsub{h}{test}(\sigma) d\sigma \right)$ and $\tsub{h}{test}(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $1$ year and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 2\,\mathrm{year}^{-1}$.
}
\figlab{test}
\end{figure}

On the other hand, consider a ``test-and-treat'' strategy in which infected individuals are identified and put to clinical care, such as antiretroviral therapy \citep{garnett2009treating, nah2017test}.
We assume that the rate of identification and removal is initially constant.
Thus, the proportion continuing to transmit declines exponentially, and the effect of intervention $\tsub{L}{test}$ increases exponentially.
However, we assume removal eventually saturates, to account for the effects of people who avoid identification, persistent treatment failures, and the possibility of rare transmission even under effective treatment
(\figref{test}A; see caption for the functional form of $\tsub{L}{test}$).
Under this assumption, the strength of intervention decreases as the proportion of early transmission increases but is always higher than the epidemic strength \citep{eaton2014proportion}.

@@

This framework provides an intuitive underpinning for the originally surprising result of \cite{eaton2014proportion}: the effectiveness of test-and-treat interventions should not depend much on the proportion of early transmission because early transmission has no effect on the estimated speed of the epidemic, and little on the estimated speed of the intervention.
